---
title: "Class 9: Candy Mini-Project"
author: "David Tin (PID: A17785446)"
format: pdf
---

## Background

Today we are taking a detour to analyze a fun dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA)

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```


## Data import
The data is all related to Halloween candy and is from the 538 website.

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r, message=FALSE}
library(tidyverse)
```


```{r}
candy %>%
  filter(fruity == T) %>%
  nrow()
```

## Favorite candy

One of the most interesting variables in the dataset is winpercent. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset (what 538 term a matchup). Higher values indicate a more popular candy.

```{r}
candy["Twix", ]$winpercent

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy |> 
  filter(row.names(candy)=="Skittles original") |> 
  select(winpercent)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy |> 
  filter(row.names(candy)=="Kit Kat") |> 
  select(winpercent)
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy |> 
  filter(row.names(candy)=="Tootsie Roll Snack Bars") |> 
  select(winpercent)
```

### skimr::skim() function

There is a useful ``skim()`` function in the **skimr** package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent mean is on a different scale. Every other variable is below 1, where win percent is at 50.32%


> Q7. What do you think a zero and one represent for the `candy$chocolate` column?

A zero and one represents whether or not a candy contains chocolate or not.

## Exploratory Analysis

A good place to start any exploratory analysis is with a histogram. You can do this most easily with the base R function hist(). Alternatively, you can use ggplot() with geom_hist(). Either works well in this case and (as always) its your choice.

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
## Base R
hist(candy$winpercent)

##ggplot2
ggplot(candy,
       aes(winpercent)) +
         geom_histogram(bins=10, fill="blue")
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values is not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

The center of the distribution is above 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
  ## chocolate candy win %
chocolatewin <- candy$winpercent[as.logical(candy$chocolate)]
  ## fruit candy win %
fruitywin <- candy$winpercent[as.logical(candy$fruity)]

## testing whether chocolate average win % > fruit candy win %
mean(chocolatewin) > mean(fruitywin)
```

On average, chocolate candy is ranked higher than fruit candy.

> Q12. Is this difference statistically significant?

```{r}
t.test(chocolatewin, fruitywin)
```

Given that the p-value is significantly below the 95% confidence interval (a=0.5), the null hypothesis is rejected, so there is a statistical significance between the win percentage between chocolate and fruit candy.

## Overall Candy Rankings

Let’s use the base R order() function together with head() to sort the whole dataset by a variable/column of interest. Or if you have been getting into the tidyverse and the dplyr package you can use the arrange() function together with head() to do the same thing and answer the following questions:

> Q13. What are the five least liked candy types in this set?

```{r}
candy %>%
  arrange(winpercent) %>%
  head(5)

```

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>%
  arrange(winpercent) %>%
  tail(5)
```

To examine more of the dataset in this vain we can make a barplot to visualize the overall rankings. As always we will use an iterative approach to building a useful visualization by getting a rough starting plot and then refining and adding useful details in a stepwise process.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy,
       (aes(winpercent, rownames(candy)))) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15, fig.width=10}
ggplot(candy,
       (aes(winpercent, 
            reorder(rownames(candy),winpercent)))) +
  geom_col()
```

### Adding useful color

Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "red"
```

Now let’s try our barplot with these colors. Note that we use `fill=my_cols` for `geom_col()`. Experiment to see what happens if you use `col=mycols`.

```{r, fig.height=15, fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is "Sixlets".

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is "Starburst".

## Taking a look at pricepercent

What about value for money? What is the best candy for the least money? One way to get at this would be to make a plot of winpercent vs the pricepercent variable. The pricepercent variable records the percentile rank of the candy’s price against all the other candies in the dataset. Lower values are less expensive and higher values are more expensive.

To this plot we will add text labels so we can more easily identify a given candy. There is a regular geom_label() that comes with ggplot2. However, as there are quite a few candies in our dataset lots of these labels will be overlapping and hard to read. To help with this we can use the geom_text_repel() function from the ggrepel package.


```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 10)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
candy %>%
  mutate(wintomoney = winpercent/pricepercent) %>%
  arrange(desc(wintomoney)) %>%
  head(1)
```

The chocolate candy, "Tootsie Roll Midgies", is the highest ranked in terms of winpercent for the least money.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy %>%
  arrange(desc(pricepercent)) %>%
  select(pricepercent, winpercent) %>%
  head(5)
```

## Exploring the Correlation Structure


```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The variables chocolate and fruity are anti-correlated.

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

I predict that chocolate and fruit will have large contributions to PC1.

## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
attributes(pca)
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate 
       other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
##library(plotly)
##ggplotly(p)
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(row.names(pca$rotation), PC1))+
  geom_col()
```

The variables that are strongly picked up by PC1 in the positive direction are fruity, pluribus, and hard. These make sense, as they have strong positive correlations shown in the earlier correlation matrix plot.

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Based on the results found, a winning candy seems to be a bar candy that is made of chocolate. From the data visualization and correlations, chocolate bar candies consistently rank high, and the PCA also demonstrates this, as the two variables of bar and chocolate shows a strong influence while being correlated with win percent.
