---
title: "Lab 18: Pertussis and the CMI-PB Project"
author: "David Tin (PID: A17785446)"
format: pdf
editor: visual
TOC: TRUE
---

## Investigating Pertussis Cases By Year

```{r}
library(datapasta)
library(tidyverse)
library(ggplot2)
```

```{r}
pertussis <- read.csv("U.S. Reported Pertussis Cases_ 1922 - 2025.csv")
cdc <- data.frame(pertussis) %>% 
  mutate(cases=as.numeric(gsub(",", "", .$Number.of.Reported.Pertussis.Cases)))
```

```{r}
ggplot(cdc) +
  aes(x=Year, y=cases) +
  geom_point() +
  geom_line(aes(group = 1)) +
  labs(x="Year", y="Number of cases", title = "Pertussis Cases by Year (1922-2025)")
```

## A tale of two vaccines (wP & aP)

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(x=Year, y=cases) +
  geom_point() +
  geom_line(aes(group = 1)) +
  labs(x="Year", y="Number of cases", title = "Pertussis Cases by Year (1922-2025)") +
  geom_vline(xintercept = 1946, color = "blue") + geom_vline(xintercept = 1996, color = "red")
```

After the introduction of the wP vaccine, the number of pertussis cases drops dramatically.

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, pertussis cases seems to rise again. A possible explanation may be the bacteria developing immunity.

## Exploring CMI-PB data

```{r}
library(jsonlite)
```
```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject, 3)
```

>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$biological_sex, subject$race)
```

>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
ap <- subject %>%
  filter(infancy_vac == "aP")
round(summary(time_length(ap$age, "years")))
```

```{r}
wp <- subject %>%
  filter(infancy_vac == "wP")
round(summary(time_length(wp$age, "years")))
```

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

I think the two groups are significantly different. The two groups clearly lie on different parts of the graph, with wP being more towards the middle than aP.

### Joining multiple tables

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
```

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

The most recent dataset has the least amount.

## Examine IgG Ab titer levels

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

FIM2/3 shows differences in the level of IgG antibody titers. This may be due to how the vaccine may target different antibodies, and in this case, Fim2/3 is targeted for immune memory.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

>Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = T) +
  facet_wrap(vars(visit)) +
  theme_bw()
```


>Q16. What do you notice about these two antigens time courses and the PT data in particular?

The OVA antigen does not change much overtime. However, the PT and FIM2/3 antibodies show a clear rise and decline over time, for both wP and aP subjects.

> Q17. Do you see any clear difference in aP vs. wP responses?

There does not seem to be a huge difference between the two types.

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

>Q18. Does this trend look similar for the 2020 dataset?

```{r}
abdata.20 <- abdata %>% filter(dataset == "2020_dataset")

abdata.20 %>% 
  filter(isotype == "IgG",  antigen == "PT", planned_day_relative_to_boost < 200) %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

Compared to the 2020 data, the trend looks similar, jumping up in MFI at the dashed, and then stabilizing after.

## Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

>Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

>Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

The expression of the gene is at its highest on visit 4.

>Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

The pattern in time does match the trend of the antibody titer, as the titers that saw significant change did so in visit 4.



